---
title: "Reproducible Research Assignment 1"
author: "B. Devan Steiner"
date: "1/10/2021"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)

```

## Loading and preprocessing the data


```{r loading and preprcessing data}
activity <- read.csv("./activity.csv")

library(ggplot2)
library(dplyr)

activity$date <- as.Date(activity$date, "%Y%m%d")
weekday <- weekdays(activity$date)
activity <- cbind(activity, weekday)
activity_nona <- subset(activity, !is.na(activity$steps))

```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day

```{r total steps}

StepsPerDay <- aggregate(activity$steps, list(activity$date), FUN=sum)
colnames(StepsPerDay) <- c("Date", "Steps")
StepsPerDay

```

2. Make a histogram of the total number of steps taken each day

```{r histogram, echo=TRUE}
g <- ggplot(StepsPerDay, aes(Steps))
g+geom_histogram(boundary=0, binwidth=2500, col="darkgreen", fill="lightgreen")+ggtitle("Histogram of steps per day")+xlab("Steps")+ylab("Frequency")+theme(plot.title = element_text(face="bold", size=12))+scale_x_continuous(breaks=seq(0,25000,2500))+scale_y_continuous(breaks=seq(0,18,2))

```

3. Calculate and report the mean and median of the total number of steps
taken each day.

mean = 10766.19
```{r mean of total steps}

mean(StepsPerDay$Steps, na.rm=TRUE)
```

median = 10765
```{r median of total steps}
median(StepsPerDay$Steps, na.rm=TRUE)
```

## What is the average daily activity pattern?
1. Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days.

```{r time series}

StepsPerTime <- aggregate(steps~interval,
                          data=activity,
                          FUN=mean,
                          na.action=na.omit)
StepsPerTime$time <- StepsPerTime$interval/100
h <- ggplot(StepsPerTime, aes(time, steps))
h+geom_line(col="brown")+ggtitle("Average steps per time interval")+xlab("Time")+ylab("Steps")+theme(plot.title = element_text(face="bold", size=12))

```


2. Which 5- minute interval, on average across all the days in the dataset, contain the maximum number of steps?

``` {r 5-min intervals}

ST <- tibble::as_tibble(StepsPerTime)
ST %>% select(time, steps) %>% filter(steps==max(ST$steps))

```

##Imputing missing values

1. Calculate and report the total number of missing values in the dataset.

```{r total number missing values}
ACT <- tibble::as_tibble(activity)
ACT %>% filter(is.na(steps)) %>% summarize(missing_values = n())
```

2. Devise a strategy for filling in all of the missing values in the dataset.

```{r filling in missing values}
activity$Complete <- ifelse(is.na(activity$steps),round(StepsPerTime$steps[match(activity$interval,StepsPerTime$interval)],0), activity$steps)

activityFull <- data.frame(steps=activity$Complete, 
                           interval=activity$interval, 
                           date=activity$date)
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

``` {r create new dataset}
StepsPerDayFull <- aggregate(activityFull$steps, list(activityFull$date), FUN=sum)
colnames(StepsPerDayFull) <- c("Date", "Steps")
```

4. Make a histogram of the total number of steps taken each day and calculate the mean and median total number of steps taken per day.

``` {histogram}
g <- ggplot(StepsPerDayFull, aes(Steps))
g+geom_histogram(boundary=0, binwidth=2500, col="darkblue", fill="lightblue")+ggtitle("Histogram of steps per day")+xlab("Steps")+ylab("Frequency")+theme(plot.title = element_text(face="bold", size=12))+scale_x_continuous(breaks=seq(0,25000,2500))+scale_y_continuous(breaks=seq(0,26,2))
```

mean = 10765.64
```{r mean}
mean(StepsPerDayFull$Steps)
```

median = 10762
```{r median}
median(StepsPerDayFull$Steps)
```

## Are there differences in activity patterns between weekdays and weekends?
1, Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
``` {r weekday vs weekend}
activityFull$RealDate <- as.Date(activityFull$date, format = "%Y-%m-%d")
activityFull$weekday <- weekdays(activityFull$RealDate)
activityFull$DayType <- ifelse(activityFull$weekday=='Saturday' | activityFull$weekday=='Sunday', 'weekend','weekday')
```

2.Two time series plot of the 5-minute interval (x) and the average number of steps taken averaged across weekday days or weekend days (y).

``` {r timeplot panel}
StepsPerTimeDT <- aggregate(steps~interval+DayType,
                            data=activityFull,
                            FUN=mean,
                            na.action=na.omit)
StepsPerTimeDT$time <- StepsPerTime$interval/100
j <- ggplot(StepsPerTimeDT, aes(time, steps))
j+geom_line(col="darkred")+ggtitle("Average steps per time interval: weekdays vs. weekends")+xlab("Time")+ylab("Steps")+theme(plot.title = element_text(face="bold", size=12))+facet_grid(DayType ~ .)
```
